---
title: "Aducanumab"
author: "Lacey Nixon"
date: "2024-04-17"
output:
  pdf_document:
    toc: true
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
subtitle: 'Preprocessing: Filtering, Normalization, Clustering'
editor_options:
  chunk_output_type: inline
---

#Setup 
##Working Directory
```{r setup}
knitr::opts_knit$set(root.dir=".")
```

##Libraries
```{r}
#install.packages('gridExtra')
library(Seurat)
library(ggplot2)
library(stringr)
library(gridExtra)
library(egg)
library(dplyr)
```

##Read in Data, Create Seurat object
```{r}
prefix <- "../counts/"
suffix <- "/outs/filtered_feature_bc_matrix.h5"


samples <- c("705A_Adu_Male", "823A_IgG_Male")
newnames <- c("adu.m.705A", "igg.m.823A")
for (i in 1:length(samples)) {
  obj <- CreateSeuratObject(Read10X_h5(paste0(prefix, samples[i], suffix)))
  assign(newnames[i], obj)
}
remove(obj)

mouse <- merge(x = adu.m.705A, y = igg.m.823A, add.cell.ids = newnames, project = "aducanumab.mice.")

remove(adu.m.705A, igg.m.823A)
gc()
```

## Variables 
Setting Thresholds for Filtering 
General ballpark: 
ncount - keeping under 25k 
nfeature - keeping under 6k 
complexity - keeping over 0.8 
```{r}
ncountmax <- 25000
ncountmin <- 0
nfeaturemax <- 6000
nfeaturemin <- 0
complexitymin <- 0.8
percent.mt.max <- 1
```


## Metadata 

###Percent MT
Filtering MT genes 
```{r}
## See features 
#View(as.matrix(rownames(mouse@assays$RNA@features)))

mouse[["percent.mt"]] <- PercentageFeatureSet(mouse, pattern = "^mt-")
mouse[["complexity"]] <- log10(mouse$nFeature_RNA) / log10(mouse$nCount_RNA)
```

"adu.m.705A", "igg.m.823A"

###Sample column 
Parentheses around part of string I want
([aduig]+\\.[mf]\\.[0-9]+[AB]) is the sample name 

```{r}
barcodes <- colnames(mouse)
sample <- str_match(barcodes, "([aduig]+\\.[mf]\\.[0-9]+[AB])_[AGTC]+\\-[0-9]")[,2]
mouse$sample <- sample 
```


#Pre-Filtering QC 
Pick favorite quality control graphs 
Visualizing 
nFeature <- unique number of genes
ncount <- total number of transcripts 
percent.mt <- percentage of mitochondrial transcripts 

##Violin Plots 
```{r}
png(file="../results/prefilteringqc/vlnplots.png", 
    width = 600, height = 350)
VlnPlot(object = mouse, 
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), 
        group.by = "sample",
        ncol = 3,
        pt.size = 0)
dev.off()
```

##Scatter plots 
Grid arrange scatter plots 
```{r}
ncount.nfeature <- ggplot(mouse@meta.data, 
                          aes(x = nCount_RNA, y = nFeature_RNA, color = percent.mt)) +
                   geom_point() + 
                   geom_hline(yintercept = 200, linetype = 'dotted') +
                   geom_hline(yintercept = 10000, linetype = 'dotted')
                   facet_wrap(~sample)

ncount.percentmt <- ggplot(mouse@meta.data, 
                          aes(x = nCount_RNA, y = percent.mt, color = percent.mt)) +
                   geom_point() + 
                   facet_wrap(~sample)

png(file="../results/prefilteringqc/prefilteringscatterplots.png", 
    width = 700, height = 450)
    grid.arrange(ncount.nfeature, ncount.percentmt,
                 nrow = 2, ncol = 1)
dev.off()
```



## Density plots 
Note - alpha is opacity 
Do for percent mt , nfeature , cell complexity
In the future - look at ribosomal and hemoglobin
Save graphs in results/prefilterngqc 

Grid Arrange - Density Plots in One Image 
```{r}
ncountdensity <-   ggplot(mouse@meta.data, 
                         aes(color = sample, 
                         x = nCount_RNA, 
                         fill = sample)) + 
                   geom_density(alpha = 0.2) + 
                   theme_classic() +
                   geom_vline(xintercept = 25000, linetype = "dashed")
ncountdensity

nfeaturedensity <-  ggplot(mouse@meta.data, 
                            aes(color = sample, 
                            x = nFeature_RNA, 
                            fill = sample)) + 
                    geom_density(alpha = 0.2) + 
                    theme_classic() +
                    geom_vline(xintercept = 6000, linetype = "dashed")
nfeaturedensity

percentmtdensity <- ggplot(mouse@meta.data, 
                           aes(color = sample, 
                           x = percent.mt, 
                           fill = sample)) + 
                    geom_density(alpha = 0.2) + 
                    theme_classic() +
                    geom_vline(xintercept = 1, linetype = "dashed")
percentmtdensity

complexitydensity <-   ggplot(mouse@meta.data, 
                              aes(color = sample, 
                              x = complexity,
                              fill = sample)) + 
                        geom_density(alpha = 0.2) + 
                        theme_classic() + 
                        geom_vline(xintercept = 0.8, linetype = "dashed")
complexitydensity


png(file="../results/prefilteringqc/prefilteringdensityplots.png", 
    width = 700, height = 450)
    grid.arrange(ncountdensity, nfeaturedensity, percentmtdensity, complexitydensity, 
                 nrow = 2, ncol = 2)

dev.off()

```


##Cells per sample 
plot of number of cells - bar graph of cells per sample (table(sample))
pre and post filtering (ncells1vsncells2)
```{r}
data <- as.data.frame(table(mouse$sample))
colnames(data) <- c("sample", "frequency")

ncells1 <- ggplot(data, 
                  aes(x = sample, y = frequency, fill = sample)) + 
                  geom_col() +
                  geom_text(aes(label = frequency), vjust = -0.5)
ncells1

png(file="../results/prefilteringqc/ncells1.png",
    width = 600, height = 350)
    ggplot(data, 
           aes(x = sample, y = frequency, fill = sample)) + 
           geom_col() +
           geom_text(aes(label = frequency), vjust = -0.5)
dev.off()
  
```
  
   
#Filtering 
Filter cells out, then filter genes out
Subset is what you are keeping 
ncount - keeping under 25k 
nfeature - keeping under 6k 
complexity - keeping over 0.8 

##Cell level Filtering
```{r}
dim(mouse)
mouse.filtered <- subset(mouse, 
                          subset = (nCount_RNA < ncountmax) & 
                            (nFeature_RNA < nfeaturemax) &
                            (complexity > complexitymin) &
                            (percent.mt < percent.mt.max))
dim(mouse.filtered) 
```


##Gene level filtering 
Keeping genes expressed in at least 10 cells 
```{r}
mouse.filtered <- JoinLayers(mouse.filtered)

counts <- GetAssayData(mouse.filtered, layer = "counts")
nonzero <- counts > 0
keep <- Matrix::rowSums(nonzero) >= 10
counts.filtered <- counts[keep,]
```

New object with filtered 
```{r}
mouse.filtered <- CreateSeuratObject(counts.filtered, meta.data = mouse.filtered@meta.data)
dim(mouse.filtered)
```


#Post-Filtering QC 
Same plots as pre-filtering QC - now use mouse.filtered object and save in results/postfilteringqc 

##Violin Plots 
```{r}
png(file="../results/postfilteringqc/vlnplots.png", 
    width = 600, height = 350)
VlnPlot(object = mouse.filtered, 
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), 
        group.by = "sample",
        ncol = 3,
        pt.size = 0)
dev.off()
```

##Scatter Plots 
```{r}
ncount.nfeature2 <- ggplot(mouse.filtered@meta.data, 
                          aes(x = nCount_RNA, y = nFeature_RNA, color = percent.mt)) +
                   geom_point() + 
                  # geom_hline(yintercept = nfeaturemax, linetype = 'dotted') +
                  # geom_hline(xintercept = ncountmax, linetype = 'dotted')
                   facet_wrap(~sample)

ncount.percentmt2 <- ggplot(mouse.filtered@meta.data, 
                          aes(x = nCount_RNA, y = percent.mt, color = percent.mt)) +
                   geom_point() + 
                   facet_wrap(~sample)

png(file="../results/postfilteringqc/postfilteringscatterplots.png", 
    width = 700, height = 450)
    grid.arrange(ncount.nfeature2, ncount.percentmt2,
                 nrow = 2, ncol = 1)
dev.off()
```

##Density Plots 
```{r}
ncountdensity2 <-   ggplot(mouse.filtered@meta.data, 
                         aes(color = sample, 
                         x = nCount_RNA, 
                         fill = sample)) + 
                   geom_density(alpha = 0.2) + 
                   theme_classic() +
                   geom_vline(xintercept = 25000, linetype = "dashed")
ncountdensity2

nfeaturedensity2 <-  ggplot(mouse.filtered@meta.data, 
                            aes(color = sample, 
                            x = nFeature_RNA, 
                            fill = sample)) + 
                    geom_density(alpha = 0.2) + 
                    theme_classic() +
                    geom_vline(xintercept = 6000, linetype = "dashed")
nfeaturedensity2

percentmtdensity2 <- ggplot(mouse.filtered@meta.data, 
                           aes(color = sample, 
                           x = percent.mt, 
                           fill = sample)) + 
                    geom_density(alpha = 0.2) + 
                    theme_classic() +
                    geom_vline(xintercept = 1, linetype = "dashed")
percentmtdensity2

complexitydensity2 <-   ggplot(mouse.filtered@meta.data, 
                              aes(color = sample, 
                              x = complexity,
                              fill = sample)) + 
                        geom_density(alpha = 0.2) + 
                        theme_classic() + 
                        geom_vline(xintercept = 0.8, linetype = "dashed")
complexitydensity2


png(file="../results/postfilteringqc/postfilteringdensityplots.png", 
    width = 700, height = 450)
    grid.arrange(ncountdensity2, nfeaturedensity2, percentmtdensity2, complexitydensity2, 
                 nrow = 2, ncol = 2)

dev.off()
```


##Cells per sample 
```{r}
data2 <- as.data.frame(table(mouse.filtered$sample))
colnames(data2) <- c("sample", "frequency")

ncells2 <- ggplot(data2, aes(x = sample, y = frequency, fill = sample)) + 
           geom_col() + 
           geom_text(aes(label = frequency), vjust = -0.5)
ncells2

png(file="../results/postfilteringqc/ncells2.png",
    width = 600, height = 350)
    ggplot(data2, 
           aes(x = sample, y = frequency, fill = sample)) + 
           geom_col()
           geom_text(aes(label = frequency), vjust = -0.5)
dev.off()


```

###Cells Pre vs Post Filtering 
```{r}
png(file="../results/postfilteringqc/ncells1vncells2.png", 
    width = 700, height = 700)
    grid.arrange(ncells1, ncells2, 
                 nrow = 2, ncol = 1)

dev.off()
```


##Top transcripts
```{r}
toptranscriptsdataframe <- data.frame(genename = rownames(mouse.filtered), 
                                      rsums = rowSums(x = mouse.filtered, slot = "counts"))
toptranscriptsdataframe <- write.table(toptranscriptsdataframe, "../results/postfilteringqc/toptranscriptsfiltered.tsv", quote = FALSE, row.names = FALSE)


```


##Boxplot log10nFeature_RNA per cell vs sample 
```{r}
png(file = "../results/postfilteringqc/log10nfeaturepercellpersample.png", 
         width = 700, height = 700)

ggplot(mouse.filtered@meta.data, aes(x = sample, y = log10(nFeature_RNA), fill = sample)) + 
geom_boxplot()

dev.off()

```


# Unwanted Variaton 
## New Function - Cell Cycle Function 
```{r}
cellcycleqc <- function(obj, species, markersPath, samplecol = "sample", outDir = NULL, verbose = TRUE){

  # Testing 
  #obj <- mouse.filtered
  #species <- "mouse"
  #markersPath <- "../refs/cell_cycle_markers.tsv"
  #samplecol <- "sample" 
  #outDir <- "../results/unwantedvariation/"
  #verbose <- TRUE
  
  # set output dir 
  output <- "./" 
  if(!is.null(outDir)) {output <- outDir}
  if(!endsWith(output, "/")) { output <- paste0(output, "/")}
 
  #Log normalization 
  if(verbose) {print("Normalizing data")}
  obj <- NormalizeData(obj)
  
  #Load cell cycle markers 
  phase.markers <- read.delim(markersPath, header = TRUE, sep = "\t") 
  phase.markers <- phase.markers[phase.markers$species==species,]
  
  #save cycle marker list 
  write.table(phase.markers, 
              paste0(output, species, "_cell_cycle_phase_markers.tsv"),
              quote = FALSE, sep = "\t", row.names = FALSE) 
  
  
  
  #subset cell cycle markers 
  ##S phase genes 
  s_genes <- phase.markers %>%
          dplyr::filter(phase == "S") %>%
          pull("gene_name") 
  
  ## G2M Genes 
  g2m_genes <- phase.markers %>%
          dplyr::filter(phase == "G2/M") %>%
          pull("gene_name")  
  
  # score cells 
  obj <- CellCycleScoring(obj, 
                          g2m.features = g2m_genes, 
                          s.features = s_genes)
  
  
  # identify most variable genes - find variable features 
  obj <- FindVariableFeatures(obj, 
                     selection.method = "vst",
                     nfeatures = 2000, 
                     verbose = FALSE)
 
  # Scale the counts 
  obj <- ScaleData(obj)
  
  # perform PCA 
  obj <- RunPCA(obj)
 
  # QC plots dim plot pc1 and pc2 
  
  qcplot1 <- DimPlot(obj,
             reduction = "pca",
             group.by= "Phase")
        
  qcplot2 <- DimPlot(obj,
             reduction = "pca",
             group.by= "Phase", 
             split.by = "Phase") 
  
 # percent phase bar graph
  meta <- dplyr::rename(obj@meta.data, phase_sample_split=sample)
  obj@meta.data <- meta 
  phase <- obj$Phase
  
  qcplot3 <- obj@meta.data %>%
    group_by(phase_sample_split, Phase) %>%
    dplyr::count() %>%
    group_by(phase_sample_split) %>%
    dplyr::mutate(percent = 100*n/sum(n)) %>%
    ungroup() %>%
    ggplot(aes(x = phase_sample_split, y = percent, fill = Phase)) +
    geom_col() +
    ggtitle("Percentage of cell phase per sample") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
    labs(x = "Sample")
  
  png(file="../results/unwantedvariation/cellcycleqcgraphs.png", 
    width = 1200, height = 400)
    
    grid.arrange(qcplot1, qcplot2, qcplot3, 
                 nrow = 1, ncol = 3)

  dev.off()

 
  # Return saves what we care about  
  return(phase)

  
  }
```

```{r}
#cellcycleqc <- function(obj, species, markersPath, samplecol = "sample", outDir = NULL, verbose = TRUE){

mouse.filtered[["phase"]] <- cellcycleqc(mouse.filtered,
                             species = "mouse",
                             markersPath = "../refs/cell_cycle_markers.tsv",
                             samplecol = "sample",
                             outDir = "../results/unwantedvariation/",
                             verbose = TRUE) 



```


## New Function - Mitochondrial Expression
```{r}
mitochondrialqc <- function(obj, samplecol = "sample", usermitocol = NULL, outDir = NULL, verbose = TRUE){

  # Testing 
  #obj <- mouse.filtered
  #samplecol <- "sample" 
  #usermitocol <- NULL
  #outDir <- "../results/unwantedvariation/"
  #verbose <- TRUE
  
  # set output dir 
  output <- "./" 
  if(!is.null(outDir)) {output <- outDir}
  if(!endsWith(output, "/")) { output <- paste0(output, "/")}
  
  # set percent.mt column 
  mito_column <- "percent.mt"
  if(!is.null(usermitocol)) {mito_column <- usermitocol}
 
  #Log normalization 
  if(verbose) {print("Normalizing data")}
  obj <- NormalizeData(obj)
  
  
  # identify most variable genes - find variable features 
  obj <- FindVariableFeatures(obj, 
                     selection.method = "vst",
                     nfeatures = 2000, 
                     verbose = FALSE)
 
  # Scale the counts 
  obj <- ScaleData(obj)
  
  # perform PCA 
  obj <- RunPCA(obj)
 
  #summary(obj@meta.data$percent.mt)
  
  #mito factor
  meta <- obj@meta.data
  first <- as.numeric(summary(meta[, mito_column])[2])
  mean <- as.numeric(summary(meta[, mito_column])[4])
  third <- as.numeric(summary(meta[, mito_column])[5])
  
  # make mito_factor column in dataframe 
  if(verbose){print("Factoring percent.mt")}
  obj[["mito_factor"]] <- cut(meta[,mito_column],
                              breaks = c(-Inf, first, mean, third, Inf),
                              labels = c("Low", "Medium", "Medium high", "High"))
  
  
 
   # QC plots dim plot pc1 pc2 
  
  qcplot1 <- DimPlot(obj,
             reduction = "pca",
             group.by= "mito_factor")
        
  qcplot2 <- DimPlot(obj,
             reduction = "pca",
             group.by= "mito_factor", 
             split.by = "mito_factor") 
  
  meta <- dplyr::rename(obj@meta.data, mito_sample_split=samplecol)
  obj@meta.data <- meta 
  
  qcplot3 <- obj@meta.data %>%
    group_by(mito_sample_split, mito_factor) %>%
    dplyr::count() %>%
    group_by(mito_sample_split) %>%
    dplyr::mutate(percent = 100*n/sum(n)) %>%
    ungroup() %>%
    ggplot(aes(x = mito_sample_split, y = percent, fill = mito_factor)) +
    geom_col() +
    ggtitle("Percentage of mito_factor per sample") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
    labs(x = "Sample")

  png(file="../results/unwantedvariation/percentmitographs.png", 
      width = 800, height = 500)
  
  plots <- list(qcplot1, qcplot2, qcplot3)
  layout <- rbind(c(2,2), c(1,3))
  grid <- grid.arrange(grobs = plots, layout_matrix=layout)
   
  dev.off()

 
  #View(obj@meta.data)
  
  #return mito vector 
  mito <- obj$mito_factor
 # phase 
 
  # Return saves what we care about  
  return(mito)
#  obj$Phase

  }
```

```{r}
#mitochondrialqc <- function(obj, samplecol = "sample", usermitocol = NULL, outDir = NULL, verbose = TRUE)
mouse.filtered[["mito_factor"]] <- mitochondrialqc(mouse.filtered,
                                                   usermitocol = "percent.mt",
                                                   samplecol = "sample",
                                                   outDir = "../results/unwantedvariation/",
                                                   verbose = TRUE) 
```


