---
title: "Aducanumab"
author: "Lacey Nixon"
date: "2024-04-17"
output:
  pdf_document:
    toc: true
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
subtitle: 'Preprocessing: Filtering, Normalization, Clustering'
editor_options:
  chunk_output_type: inline
---

#Setup 
##Working Directory
```{r setup}
knitr::opts_knit$set(root.dir=".")
```

##Libraries
```{r}
#install.packages('gridExtra')
library(Seurat)
library(ggplot2)
library(stringr)
library(gridExtra)
library(egg)
library(dplyr)
```

#Read in Data 

```{r}
mouse.unannotated <- readRDS("../robjects/mouse.unannotated.RDS")
```


#Variables 


#Auto find markers 
```{r}
# Set seurat clusters 
mouse.unannotated$seurat_clusters <- mouse.unannotated$SCT_snn_res.0.2

DefaultAssay(mouse.unannotated) <- "RNA"
mouse.unannotated <- NormalizeData(mouse.unannotated)
Idents(mouse.unannotated) <- "seurat_clusters"


#Kept getting error - ask Kennedi 
markers <- SeuratWrappers::RunPrestoAll(
  object = mouse.unannotated, 
  assay = "RNA", 
  slot = "counts", 
  only.pos = FALSE
)

# This one takes less time 
#markers <- FindAllMarkers(object = mouse.unannotated,
#                          only.pos = FALSE,
#                          logfc.threshold = 0.25)

#Filtered adjusted pvalues 
markers <- markers[markers$p_val_adj > 0.01,]



#Conserved markers by cluster 
cluster0_conserved_markers <- FindConservedMarkers(mouse.unannotated,
                              ident.1 = 0,
                              grouping.var = "sample",
                              only.pos = TRUE,
                              logfc.threshold = 0.25)
cluster1_conserved_markers <- FindConservedMarkers(mouse.unannotated,
                              ident.1 = 1,
                              grouping.var = "sample",
                              only.pos = TRUE,
                              logfc.threshold = 0.25)
cluster2_conserved_markers <- FindConservedMarkers(mouse.unannotated,
                              ident.1 = 2,
                              grouping.var = "sample",
                              only.pos = TRUE,
                              logfc.threshold = 0.25)
cluster3_conserved_markers <- FindConservedMarkers(mouse.unannotated,
                              ident.1 = 3,
                              grouping.var = "sample",
                              only.pos = TRUE,
                              logfc.threshold = 0.25)
cluster4_conserved_markers <- FindConservedMarkers(mouse.unannotated,
                              ident.1 = 4,
                              grouping.var = "sample",
                              only.pos = TRUE,
                              logfc.threshold = 0.25)
cluster5_conserved_markers <- FindConservedMarkers(mouse.unannotated,
                              ident.1 = 5,
                              grouping.var = "sample",
                              only.pos = TRUE,
                              logfc.threshold = 0.25)
cluster5_conserved_markers <- FindConservedMarkers(mouse.unannotated,
                              ident.1 = 5,
                              grouping.var = "sample",
                              only.pos = TRUE,
                              logfc.threshold = 0.25)
cluster6_conserved_markers <- FindConservedMarkers(mouse.unannotated,
                              ident.1 = 6,
                              grouping.var = "sample",
                              only.pos = TRUE,
                              logfc.threshold = 0.25)
cluster7_conserved_markers <- FindConservedMarkers(mouse.unannotated,
                              ident.1 = 7,
                              grouping.var = "sample",
                              only.pos = TRUE,
                              logfc.threshold = 0.25)
cluster8_conserved_markers <- FindConservedMarkers(mouse.unannotated,
                              ident.1 = 8,
                              grouping.var = "sample",
                              only.pos = TRUE,
                              logfc.threshold = 0.25)
cluster9_conserved_markers <- FindConservedMarkers(mouse.unannotated,
                              ident.1 = 9,
                              grouping.var = "sample",
                              only.pos = TRUE,
                              logfc.threshold = 0.25)
cluster10_conserved_markers <- FindConservedMarkers(mouse.unannotated,
                              ident.1 = 10,
                              grouping.var = "sample",
                              only.pos = TRUE,
                              logfc.threshold = 0.25)
cluster11_conserved_markers <- FindConservedMarkers(mouse.unannotated,
                              ident.1 = 11,
                              grouping.var = "sample",
                              only.pos = TRUE,
                              logfc.threshold = 0.25)
cluster12_conserved_markers <- FindConservedMarkers(mouse.unannotated,
                              ident.1 = 12,
                              grouping.var = "sample",
                              only.pos = TRUE,
                              logfc.threshold = 0.25)
cluster13_conserved_markers <- FindConservedMarkers(mouse.unannotated,
                              ident.1 = 13,
                              grouping.var = "sample",
                              only.pos = TRUE,
                              logfc.threshold = 0.25)

```
```{r}
# Graphs - wanted to look at some of the markers  

#Astrocyte
FeaturePlot(object = mouse.unannotated, 
                        features = c("Gja1"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) + 
  scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red"))

# Astrocyte
FeaturePlot(object = mouse.unannotated, 
                        features = c("Slc1a3"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) +
  scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red"))

#Astrocyte 
FeaturePlot(object = mouse.unannotated, 
                        features = c("Slc1a2"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) +
  scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red"))

#Astrocyte
FeaturePlot(object = mouse.unannotated, 
                        features = c("Atp1a2"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) +
  scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red"))


# Microglia 
FeaturePlot(object = mouse.unannotated, 
                        features = c( "C1qb"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) +
  scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red"))

FeaturePlot(object = mouse.unannotated, 
                        features = c( "C1qc"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) +
  scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red"))

#Microglia
FeaturePlot(object = mouse.unannotated, 
                        features = c( "Cx3cr1"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) +
  scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red"))

#Microglia
FeaturePlot(object = mouse.unannotated, 
                        features = c( "Csf1r"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) +
  scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red"))

# Endothelial 
FeaturePlot(object = mouse.unannotated, 
                        features = c("Flt1"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) +
  scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red"))

# Fibroblast 
FeaturePlot(object = mouse.unannotated, 
                        features = c("Dcn"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) +
  scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red"))

# Polydendrocyte  
FeaturePlot(object = mouse.unannotated, 
                        features = c("Pdgfra"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) +
  scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red"))

# Polydendrocyte
FeaturePlot(object = mouse.unannotated, 
                        features = c("Cspg4"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) +
  scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red"))

# Polydendrocyte
FeaturePlot(object = mouse.unannotated, 
                        features = c("Vcan"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) +
  scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red"))


# Neuron/fibroblast/microglia - came up all three 
FeaturePlot(object = mouse.unannotated, 
                        features = c("Sh3rf2"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) +
  scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red"))

# Neuron/fibroblast/microglia - came up all three 
FeaturePlot(object = mouse.unannotated, 
                        features = c("Rgs9"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) +
  scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red"))

# Neuron/fibroblast/microglia - came up all three 
FeaturePlot(object = mouse.unannotated, 
                        features = c("Rarb"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) +
  scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red"))

#Neuron/Microglia 
FeaturePlot(object = mouse.unannotated, 
                        features = c("Actn2"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) +
  scale_colour_gradientn(colours = c("blue", "lightblue", "yellow", "orange", "red"))

# Flt1 Cluster 8
# Ptprb Cluster 8
# Vwf Cluster 8 
# Prox1 Maybe Cluster 5, 6, 11, 13
# Col1a1 Maybe Cluster 3, 
# Csf1r Cluster 7
# C1qa (C1qb, C1qc) Cluster 7, maybe 12 
# Kit Cluster 6 (maybe) 
# Notch3 Cluster 8
# Rgs5 Cluster 8 
# Plp1 Cluster 5 
# Cdh19 Cluster 3, Cluster 5 
# Ms4a4b, Cd3g, Cd3d, Cd3e, Nkg7 - Cluster 12 



# Other 
FeaturePlot(object = mouse.unannotated, 
                        features = c("Lef1"),
                         order = TRUE,
                         min.cutoff = 'q10', 
                         label = TRUE,
            repel = TRUE) +
  scale_colour_gradientn(colours = c("white", "lightblue", "yellow", "orange", "red"))

```



database: dropVIz - single cell database of mouse brain annotation markers 
neurons - 50% of cell type - annotate neuron, don't get into subsets 
astrocytes, microglia, endothelial cells, fibroblasts, pericytes/smooth muscle cells (SMCs), 

astrocytes: Gja1, Gli3, Bmpr1b, Clu
microglia: C1qb, Cx3Cr1, Mrc1
endothelial: Flt1, Lef1
fibroblast: Dcn 
pericytes/mural: Rgs5, Acta2
neuron: Syt1, Bcl6, Parm1 
interneuron: Gad2, Pvalb

```{r}
# Astrocytes
VlnPlot(object = mouse.unannotated,
        features = c("Gja1", "Gli3","Bmpr1b"))

VlnPlot(object = mouse.unannotated,
        features = c("Slc1a3", "Atp1a2", "Nwd1"))

# Microglia 
VlnPlot(object = mouse.unannotated,
        features = c("C1qb", "Cx3cr1", "Mrc1"))

VlnPlot(object = mouse.unannotated,
        features = c("Csf1r", "Ly86", "C1qc"))

# Endothelial 
VlnPlot(object = mouse.unannotated,
        features = c("Flt1", "Lef1", "Rai14"))

# Fibroblast 
VlnPlot(object = mouse.unannotated,
        features = c("Dcn"))

# Pericytes/Mural 
VlnPlot(object = mouse.unannotated,
        features = c("Rgs5", "Acta2"))

# Neuron 
VlnPlot(object = mouse.unannotated, 
        features = c("Syt1", "Bcl6", "Parm1"))

# Interneuron
VlnPlot(object = mouse.unannotated,
        features = c("Gad2", "Pvalb"), stack = TRUE, flip = TRUE, group.by = "seurat_clusters", split.by = "seurat_clusters")


# All 
VlnPlot(object = mouse.unannotated,
        features = c("Gja1", "Gli3", "Bmpr1b", "Clu"), stack = TRUE, flip = TRUE, group.by = "seurat_clusters", split.by = "seurat_clusters")
VlnPlot(object = mouse.unannotated,
        features = c("C1qb", "Cx3cr1", "Mrc1"), stack = TRUE, flip = TRUE, group.by = "seurat_clusters", split.by = "seurat_clusters")
VlnPlot(object = mouse.unannotated,
        features = c("Flt1", "Lef1", "Ptprb", "Ly6c1", "Vwf"), stack = TRUE, flip = TRUE, group.by = "seurat_clusters", split.by = "seurat_clusters")
VlnPlot(object = mouse.unannotated,
        features = c("Dcn", "Col1a1", "Col1a2"), stack = TRUE, flip = TRUE, group.by = "seurat_clusters", split.by = "seurat_clusters")
VlnPlot(object = mouse.unannotated,
        features = c("Notch3", "Rgs5", "Acta2"), stack = TRUE, flip = TRUE, group.by = "seurat_clusters", split.by = "seurat_clusters")
VlnPlot(object = mouse.unannotated,
        features = c("Syt1", "Bcl6", "Parm1", "Gad2"), stack = TRUE, flip = TRUE, group.by = "seurat_clusters", split.by = "seurat_clusters")
VlnPlot(object = mouse.unannotated,
        features = c("Plp1", "Cdh19", "Mpz"), stack = TRUE, flip = TRUE, group.by = "seurat_clusters", split.by = "seurat_clusters")


# Flt1 Cluster 8
# Ptprb Cluster 8
# Vwf Cluster 8 
# Prox1 Maybe Cluster 5, 6, 11, 13
# Col1a1 Maybe Cluster 3, 
# Csf1r Cluster 7
# C1qa (C1qb, C1qc) Cluster 7, maybe 12 
# Kit Cluster 6 (maybe) 
# Notch3 Cluster 8
# Rgs5 Cluster 8 
# Plp1 Cluster 5 
# Cdh19 Cluster 3, Cluster 5 
# Ms4a4b, Cd3g, Cd3d, Cd3e, Nkg7 - Cluster 12 
```

Start annotating clusters 
ident function 

Get final UMAP, annotate 

